Icicle: a single-pass query language

### 2016-02-03 2016icicle/section/01-Introduction.tex
Add notes

For feature generation for machine learning, our queries have some quite specific restrictions.
Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.
Each query is implicitly partitioned by the entity, the company code in this case, and performs an aggregate over that company's entries, sorted by date.
In this way, our query language is actually a streaming dataflow language with only one input stream and clock.
By making use of these restrictions, we are able to generate more efficient code than a general query language.

>>>>

\ben{What is a concrete example of an SQL query where it's not immediatly obvious how many passes through the data it needs? We've claimed that for some queries the number of required passes is not ``immediately obvious'', and now need to back that up}

For feature generation for machine learning, our queries have some quite specific restrictions. \ben{The following describes properties that that queries and data have intrinsically, not \emph{restrictions} placed on them}.
Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.
Each query is implicitly partitioned by the entity, the company code in this case, and performs an aggregate over that company's entries, sorted by date.
In this way, our query language is actually a streaming dataflow language with only one input stream and clock.
By making use of these restrictions, we are able to generate more efficient code than a general query language.

### 2016-02-03 2016icicle/section/02-Source.tex
type stuff
Add notes

In this section we describe the grammar and evaluation semantics of the Icicle language.

Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
We define the stocks table as follows, leaving out the company code (entity identifier) and date.

\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$@table@~\mi{stocks}$    \\
$\{$ \> $\mi{open}$ \> $:~\NN$ \\
$~,$ \> $\mi{close}$ \> $:~\NN$ \\
$\}$
\end{tabbing}

This defines the columns as natural numbers, and allows them to be referred to by name.
In the definitions and queries below, any reference to these names means the column itself, with the entire data for a given entity.
This is annotated with the mode @Element@, meaning that it exists for every element of the input.

@@

First we define a function to compute the sum of some input data.
While traditionally sum would take just an input stream of numbers, we separate this into two distinct parameters.
The first parameter is the clock for the input stream, which defines the parts of the input stream to sample.
This can be thought of as a stream of boolean flags, where true denotes that the input stream should be sampled at this time.
The second parameter is the actual value to sample at the clock.
In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.

By separating the clock from the element values, we guarantee that any references to elements denote the same point in the stream, and can always be used together.
For example, computing the difference between the two fields above ($\mi{open} - \mi{close}$) refers to the open and close fields at the same point.
The clock is used when performing a fold, and determine at which points to sample the elements and update the accumulator.
This is in constrast to traditional streaming dataflow languages, where the element data and the clock are intertwined.
This requires clock analysis\CITE to tell whether two element streams can be used together with a bounded buffer.

We denote clocks in brackets as they are a different universe to normal expressions.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{sum}[c]~v$    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
\end{tabbing}

The result of $\mi{sum}$ above has type $\Agg{\NN}$.
Aggregate types denote that their value is only available after the entire input stream has been seen.
By restricting the bodies of folds to be element types, and their output to be aggregate types, it means no reference to the result of folds can be made in the body of a fold.

We can define a query that computes the sum of each company's open prices.
Query definitions are similar to function definitions, except they only take one argument: the clock describing all available input.
They also must return aggregates, as an element query could be unbounded in size.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@query@ 
$\mi{opensum}[c]$    \\
\> $=$  \> $\mi{sum}[c]~\mi{open}$. \\
\end{tabbing}

@@

\begin{code}
CODE, OPEN, CLOSE, DATE,        OPENSUM
ABC,    10,    12, 2015-01-01,       10
IAG,    11,    12, 2015-01-01
IAG,    12,    10, 2015-01-02,       23
\end{code}

Next, we can define functions $\mi{count}$ and $\mi{mean}$ in terms of $\mi{sum}$.
Count passes the argument $1$ to sum, and this is interpreted as an element containing all ones.
The input clocks are passed along as-is.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{count}[c]$                                     \\
 \> $=$  \> $\mi{sum}[c]~1$.                           \\
                                                    \\
@function@ 
$\mi{mean}[c]~v$                                       \\
 \> $=$  \> $\mi{sum}[c]~v~/~\mi{count}[c]$.              \\
\end{tabbing}

@@

In the example below we have shown the result of the element expression $\mi{close}~>~\mi{open}$ for clarity only; this is not included in the output of the $\mi{gap}$ query.
\begin{code}
CODE, OPEN, CLOSE, CLOSE > OPEN, GAP
ABC,    10,    12,            T,   1
IAG,    11,    12,            T
IAG,    12,    10,            F,   1
\end{code}

@@

\input{figures/Source-Grammar.tex}

In figure~\ref{fig:source:eval} we describe the bigstep evaluation semantics of the language.
The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.
The evaluation works on a single entity's data at a time, and the column values, such as $\mi{open}$ and $\mi{close}$ in the examples above, must be substituted into the functions and expression beforehand.
The clock is initially set to true for all values in the input stream.

@@

In the evaluation rules we make the distinction between @Scalar@ and @Elements@ explicit, but in the source grammar we omit the @Scalar@ qualifier, and do not allow @Elements@ as values, except when substituted in for evaluation.
For example, the definition of $\mi{sum}$ would actually have $@Scalar@~0$ as its initialiser.
Scalars can also be thought of as a repeated stream of the same value, of the same length as the input stream.

>>>>

In this section we describe the grammar and evaluation semantics of the Icicle language.
\ben{The running example here should be driven by solving the stated problem introduced in the introduction.}

Icicle queries execute over a single table, which must at least contain an entity identifier and a date. \ben{Why does the fact that it's a date matter? Could we also use any other data type that has an ordering?}

Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
We define the stocks table as follows, leaving out the company code (entity identifier) and date.

\ben{If this table really has a data field as well then why do we leave it implicit? That's a bit confusing}.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
$@table@~\mi{stocks}$    \\
$\{$ \> $\mi{open}$ \> $:~\NN$ \\
$~,$ \> $\mi{close}$ \> $:~\NN$ \\
$\}$
\end{tabbing}

This defines the columns as natural numbers, and allows them to be referred to by name.
In the definitions and queries below, any reference to these names means the column itself, with the entire data for a given entity.
This is annotated with the mode @Element@, meaning that it exists for every element of the input.
\ben{What is annotated with the mode? The column itself or the type of the column?}

@@

First we define a function to compute the sum of some input data.
While traditionally sum would take just an input stream of numbers, we separate this into two distinct parameters.
The first parameter is the clock for the input stream, which defines the parts of the input stream to sample.
This can be thought of as a stream of boolean flags, where true denotes that the input stream should be sampled at this time.
The second parameter is the actual value to sample at the clock.
In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.
\ben{Maybe hold off the description of @Element@ and @Aggregate@ until later so we can introduce them both at the same time. At this point in the text it's not clear why we need the modal types.}

By separating the clock from the element values, we guarantee that any references to elements denote the same point in the stream, and can always be used together.
For example, computing the difference between the two fields above ($\mi{open} - \mi{close}$) refers to the open and close fields at the same point.
The clock is used when performing a fold, and determine at which points to sample the elements and update the accumulator.
This is in constrast to traditional streaming dataflow languages, where the element data and the clock are intertwined.
This requires clock analysis\CITE to tell whether two element streams can be used together with a bounded buffer. \ben{The need for clock analysis needs to be backup up with an example. Readers that aren't familiar with the semantics of data flow languages won't understand the phrase ``the element and the clock are intertwined''}

We denote clocks in brackets as they are a different universe to normal expressions.
\ben{Explain the syntax. What are $v$ $s$ and @then@ about?} \ben{Mention that $s$ is a stream. What is its type? Where are $c$ and $s$ bound?}

\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{sum}[c]~(v~:~\NN)$    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
\end{tabbing}

The result of $\mi{sum}$ above has type $\Agg{\NN}$.

\ben{Does the concrete syntax support this annotation, eg:}
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{sum}[c]~(v~:~\NN) : \Agg{\NN}$    \\
\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
\end{tabbing}

Aggregate types denote that their value is only available after the entire input stream has been seen. 
By restricting the bodies of folds to be element types, and their output to be aggregate types, it means no reference to the result of folds can be made in the body of a fold. \ben{Give a counter-example of what can go wrong it we don't have this restriction}

We can define a query that computes the sum of each company's open prices.
Query definitions are similar to function definitions, except they only take one argument: the clock describing all available input.
They also must return aggregates, as an element query could be unbounded in size.
\ben{The above describes the syntax of the query, but what is the essential reason for defining functions and queries with different language constructs? Could we just say that a query is a function with some particular restrictions?}
\ben{Can we also provide the aggregate result type on the query definition itself?}

\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@query@ 
$\mi{opensum}[c]$    \\
\> $=$  \> $\mi{sum}[c]~\mi{open}$. \\
\end{tabbing}

@@

\begin{code}
CODE, OPEN, CLOSE, DATE,        OPENSUM
ABC,    10,    12, 2015-01-01,       10
IAG,    11,    12, 2015-01-01
IAG,    12,    10, 2015-01-02,       23
\end{code}
\ben{The fact that there is also an implicit group-by operation being performed is not made explicit in the query code. The operation being performed above is a segmented fold over the data, rather than a complete fold. Suppose there was also a @VOLUME@ field for the number of shares traded per day for each stock, and we wanted to compute the total volume of all shares traded on every day we have on record? How would we express that in the query language?}

Next, we can define functions $\mi{count}$ and $\mi{mean}$ in terms of $\mi{sum}$.
Count passes the argument $1$ to sum, and this is interpreted as an element containing all ones.

\ben{It's interpreted as an element containing all ones, but not a \emph{stream} containing all ones. The fact that the input data is implicitly partitioned by the key need to be stated up front, at it seems like a key part of the language design. Suppose we go back to the @VOLUME@ example and want to broadcast the value @1@ to \emph{all} the rows in the entire data set. How would we do that?}

The input clocks are passed along as-is.
\begin{tabbing}
MM \= MM \= MMMMMMMMMMM \= \kill
@function@ 
$\mi{count}[c]$                                     \\
 \> $=$  \> $\mi{sum}[c]~1$.                           \\
                                                    \\
@function@ 
$\mi{mean}[c]~(v~:~\NN)$                                       \\
 \> $=$  \> $\mi{sum}[c]~v~/~\mi{count}[c]$.              \\
\end{tabbing}

@@

In the example below we have shown the result of the element expression $\mi{close}~>~\mi{open}$ for clarity only; this is not included in the output of the $\mi{gap}$ query.

\ben{That's not what \emph{gap} means. A stock price has ``gapped up'' when it closes at some price $c$ on day 0, opens at some price $o$ on day 1, and $o > c$. The point is that there is a price discontinuity between trading days, and there was no opportunity to buy and sell at any price between $c$ and $o$.}

\begin{code}
CODE, OPEN, CLOSE, CLOSE > OPEN, GAP
ABC,    10,    12,            T,   1
IAG,    11,    12,            T
IAG,    12,    10,            F,   1
\end{code}

@@

\input{figures/Source-Grammar.tex}
\ben{Should the set of types also include @Date@? In the first example the table also contained dates}.

In figure~\ref{fig:source:eval} we describe the bigstep evaluation semantics of the language.

The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.

\ben{The previous section states that ``all function definitions can be inlined into the user query''. If this is true then why not always inline them? Why does the evaluation semantics need an environment containing a separate set of functions?}

The evaluation works on a single entity's data at a time, and the column values, such as $\mi{open}$ and $\mi{close}$ in the examples above, must be substituted into the functions and expression beforehand. \ben{What is a ``column value''?}
The clock is initially set to true for all values in the input stream.

@@

\ben{The syntax @Elements@ looks similar to @Element@, which was a type. Can we usual list or set syntax to present these?} \ben{I think it would look cleaner to have an overall construct to express the combination of input data expression to execute. The examples feel like they're setting up the big-step judgement in a particular way to get around the lack of an overall program/module construct.}

\ben{I don't follow the following paragraph.} In the evaluation rules we make the distinction between @Scalar@ and @Elements@ explicit, but in the source grammar we omit the @Scalar@ qualifier, and do not allow @Elements@ as values, except when substituted in for evaluation.
For example, the definition of $\mi{sum}$ would actually have $@Scalar@~0$ as its initialiser.
Scalars can also be thought of as a repeated stream of the same value, of the same length as the input stream.

### 2016-02-03 2016icicle/section/03-Types.tex
type stuff
Add notes

The @Pure@ modality is actually the absence of a mode; this is simply made explicit for clarity.

@@

There is also no way to take an @Aggregate@ and convert it to a @Pure@ or an @Element@, except for the @scan@ primitive which explicitly uses a fold's intermediate value, rather than the value of the entire pass.

>>>>

The @Pure@ modality is actually the absence of a mode; this is simply made explicit for clarity.
\ben{Standard modal logics use $P$, $\Diamond\,P$ and $\Box\,P$. A proposition $P$ which is universaly true does not have an associated mode. It's a similar story with @IO Int@ and @Int@ in Haskell.}

@@

There is also no way to take an @Aggregate@ and convert it to a @Pure@ or an @Element@.
